---
title: "Data Challenge #4: National Parks"
author: "Ceyda Durmaz"
date: "11/11/2020"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
editor_options: 
  chunk_output_type: console
---

Link to GitHub Repository: https://github.com/cdurmaz569/Data_Science_I_Homework/tree/master/Homework_4

# Load libraries, load data, and set options
```{r warning=FALSE, message=FALSE}
# load libraries 
libraries = c("tidytuesdayR", "knitr", "kableExtra", "data.table", "tidyverse",
              "ggplot2", "outliers", "rlang", "RColorBrewer")
library.check <- lapply(libraries, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# Get coffee data from tidytuesdayR package
tues_data <- tt_load('2019-09-17')
park_visits <- tues_data$national_parks
state_pop <- tues_data$state_pop

# Disable scientific notation
options(scipen = 999)

# Set theme
theme_set(theme_classic())

# Set seed
set.seed(18)
```

# Functions
```{r warning=FALSE, message=FALSE}
#' Plot Line Plot of Visitors over Time 
#'
#' Plots number of park visitors over time for parks. Visitor counts can be 
#' either raw or normalized by state population 
#' 
#' @param df Dataframe or list with columns 'year', 'parkname', 'visitors',
#' and 'visitors_norm'
#' @param norm Boolean indicating whether counts to plot are normalized
#'
#' @return Line plot of specified counts over time for each park 
plot_visitors <- function(df, norm = TRUE){
  
  # set labels
  if (norm) {
    y_var = "visitors_norm"
    counts = "Normalized"
    y_units = "Visitors/State Population"
  } else{
    y_var = "visitors"
    counts = "Raw"
    y_units = "Raw Number of Visitors"
  }
  
  # Visualize line plot
  ggplot(df, aes(x = year)) +
    geom_line(aes_string(y = y_var, col = "parkname")) +
    scale_color_brewer(palette = "Dark2") +
    labs(
      title = paste0(counts, " Counts of Visitors Over Time"),
      x = "Year",
      y = paste0("Number of Visitors (", y_units, ")"),
      col = "Park Name"
    )
  
}

#' Simulate Number of Visitors
#'
#' Runs simulation for number of visitors at a park for the year 1942
#' 
#' @param df Dataframe or list with columns 'year', 'parkname', 'visitors',
#' and 'visitors_norm'
#' @param park_name String of park name
#' @param n_sim Numeric of desired number of simulations. Default = 1000
#' @param prob Double of probability of visiting park
#'
#' @return Histogram of distribution for number of park visitors in 1942
sim_visits <- function(df, park_name, n_sim = 1000, prob = 0.5){
  
  # filter df to keep years 1941 and 1942 for specified park
  df_filt <- df %>% 
    filter(parkname == park_name) %>%
    filter(year >= 1941 & year <= 1942)
  
  # get sum of number of visitors in 1941 and 1942
  sum_visitors <- sum(df_filt$visitors)
  
  # get observed counts for 1942 only
  visitors_1942 <- filter(df_filt, year == 1942)$visitors 
  
  # simulate number of visitors
  sim_data <- data.frame(visitors = rbinom(n_sim, sum_visitors, prob))
  
  # Visualize simulations as histogram
  ggplot(sim_data, aes(x = visitors)) + 
    geom_histogram(fill = "steelblue4", binwidth = 10) + 
    geom_hline(yintercept = 0, col = "black") + 
    geom_vline(xintercept = visitors_1942, col = "red", size = 2) +
    labs(
      title = paste0(n_sim,
                     " Simulations of Visitors in 1942 at ",
                     park_name),
      x = "Number of Visitors",
      y = "Simulation Count",
      legend.position = "none"
    ) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
}
```

# Part 1: Trend in Park Visitors for Crater Lake, Joshua Tree, and Zion
```{r warning=FALSE, message=FALSE}
# filter by states
state_pop <- state_pop %>%
  filter(state == "CA" | state == "OR" | state == "UT")

# filter by parks and add state population data 
three_parks <- park_visits %>%
  filter(parkname == "Crater Lake" | 
         parkname == "Joshua Tree" | 
         parkname == "Zion") %>%
  select(year, state, parkname, visitors) %>% # clean-up columns
  filter(year != "Total") %>% # discard rows where year == "Total"
  mutate(year = as.numeric(year)) %>% # convert years from char to numeric
  filter(year > 1940 & year < 2016) %>% # keep years 1941-2015 
  left_join(state_pop, by = c("year", "state")) %>%
  mutate(visitors_norm = visitors/pop)

# create list of args for plot_visitors
plot_args <- list(df = list(three_parks, three_parks),
                  norm = c(FALSE, TRUE))

# plot visitors over time by raw count and normalized counts
pmap(plot_args, plot_visitors)
```

# Part 2: Simulation of visitors in 1942
```{r warning=FALSE, message=FALSE}
# create list of args for sim_visits
sim_args <- list(df = list(three_parks, three_parks, three_parks),
                 park_name = c("Joshua Tree", "Crater Lake", "Zion"))

# plot distributions of simulated number of visitors 
pmap(sim_args, sim_visits)
```

# Part 3: Summary
The trends visualized by line plots in part I show that raw counts can be 
misleading. The raw counts show that Joshua Tree has the second highest
increase, along with a steady increase, in visitors from 1940 to 2015. However, 
the normalized counts show that Joshua Tree appears relatively unchanged over time compared to the other parks. Further, the normalized counts reveal some trends 
in the data not shown visualized by raw counts. The simulation used to examine the drop in visitors from 1941 to 1942 across all three parks were consistent with the trends of the normalized line plot, where Crater Lake's and Zion Parks' observed visitors in 1942 were strikingly lower than those projected by the simulation compared to Joshua Tree. 

